Measuring the double-drift illusion with hand tracking

### 2021-08-04 Manuscript.Rmd
lots of edits to the manuscript

In our second experiment, we look at the spatial and temporal limits of the illusion. At some point, the illusion stops accumulating. The accumulation can be reset by introducing a temporal break [@Lisi2015] or by distraction attention [@Nakayama2020], but it may also reset spontaneously once the accumulation has gone on too long or too far. Informally, some observers have reported that the gabor’s path may saturate so that it  remains parallel to, but offset from the true path. Others report that the position resets back toward its physical location whereupon accumulation begins again. The ability to record trajectories of the perceived position will then allow us to study spontaneous reset outcomes.

@@

If $K$ is 0, then there is no illusion and the perceived direction matches the external motion direction. However, if $K$ is 1, then the internal motion has equal contribution with the external motion in determining the perceived direction. For comparison with other work, we will find the value of $K$ that best describes the strength of the illusion in both of our two experiments.

@@

```{r fig2, echo=FALSE, message=FALSE, fig.width=8, fig.height=5, fig.cap="\\label{fig:fig2}**Figure 2:** Online Tracking Data. **Left:** Individual trajectories. Note that there are only a few obvious discontinuities (highlighted in red). In general, the traced paths showed rough linearity. **Top right:** Average heading of the middle 1 _s_ of each pass. Distance from origin is time, angle is the heading angle. Dark blue lines the average heading, in light blue a polar heatmap with the distribution of heading. **Bottom right:** The strength of the illusion in degrees angle deviation from the physical path. In red are the average illusions strengths for 0 _cps_, ±1 _cps_ and ±3 _cps_, and in light blue the data for the 4 individual participants. Gray dashed line: prediction of the vector combination model of Equation 1 with $K=1$. Black line: fit with $K=0.81$ as in Cavanagh & Tse (2019), $K=0.81$. Purple: best fit to our data, $K=0.72$."}
plotBoundedTracking()
```

@@

The individual trajectories reveal some points that are potential spontaneous resets that we try to extract. Using a smooth spline interpolated version of the trajectory we detect right-left direction changes, based on sign changes. Since the initial movement is often jittery, we only take direction changes further than ~0.39 cm ( ~0.23 _dva_ ) away from the starting point. The first (if any) of the remaining direction changes, are then taken as reset points. By first averaging the x and y coordinates (in _cm_) of reset points in all 9 participants and 6 stimulus conditions, we obtain 54 average reset points for further analyses and modelling. Both the horizontal and vertical (straight ahead) distance of the reset point from the origin of the trajectory are converted to centimeters, and trajectories are scaled in both dimensions so that vertical distance from the origin is 13.5 cm. The vertical component can then be converted to time as 1 _cm_ = 3.375 _s_ in the 4 _s_ condition and 1 _cm_ = 4.5 _s_ in the 3 _s_ condition.

@@

```{r fig4, echo=FALSE, message=FALSE, fig.width=8, fig.height=3.5, fig.cap="\\label{fig:fig4}**Figure 4:** Re-tracing Data. **Left:** One participant's trajectories with very clear 'return' resets at the purple dots. (Purple lines: trajectories up to resets, gray lines: trajectories after resets, blue lines: trajectories that had no resets in the re-tracing.) Other participants' data were not this clear. **Middle:** Average reset point locations for all 9 participants in all 6 conditions (54 points) with standard deviations. (Gray arrow: real trajectory, blue line: spatial average, red line: temporal average.) **Right:** Illusion strength as in Figure 2 (bottom right); plotted against predicted strength from Eq.1: the arctangent of the ratio of internal to external velocities. Blue dots from 3 s passes, red from 4 s passes, open circles are group averages. Predictions from Eq. 1: Gray dashed line for $K=1$, black line; $K=0.81$, purple line; best fit here at $K=0.51$."}
#plotExampleData()
plotData()
```

On average, the algorithm detects resets in ~75% of trials. For three participants, only about half the trials showed resets (42%, 44% and 58%). For the remaining six participants the algorithm detected resets in 72%-99% of trials, with an average of 88%. The minimum number of resets in each of the six conditions for any participant is 7 (out of 24 trials), with an average of ~17. Since participants experiencing a "hit-the-wall reset" (Fig. 3) as opposed to a "jump" reset, might not always display a detectable direction change in their paths, 50% of trials with resets seems reasonable. Either type of perceived resets appears valid, and the average reset positions still indicates some limit to the illusion. Therefore, we kept all participants and resets in our analysis (reset points in Fig. 4, middle).

We compare the strength of the illusion as measured with half-way points (see Fig 4, right), with an earlier study [@Cavanagh2019]. A line through the origin with slope 0.51 ($K$ from Eq. 1) predicts illusion strength fairly well in these data, which is weaker than the $K$=0.81 found by Cavanagh & Tse [-@Cavanagh2019] and lower than the $K$=0.72 found in Experiment 1 with online tracking. We can also see that illusion strength measured with re-tracing is more variable. While the illusion is less strong and more variable, it is nevertheless there and depends on the relationship between internal and external speed as in previous studies.

@@

Now we look at the distribution of reset points, and whether they can be explained by a temporal or spatial limit. In Figure 4 (middle), we can see that the reset points vary more along the y-axis, than along the x-axis. That is, it looks like predicting reset points from a single X coordinate (or spatial limit) than predicing reset points from a given time.

```{r fig5, echo=FALSE, message=FALSE, fig.width=7, fig.height=5.5, fig.cap="\\label{fig:fig5}**Figure 5:** Reset Models. Model limit shown here are given by weighted bootstrapping from all participants, lines are median parameter values, shaded areas the 95% confidence interval of the limits. Gray dots are the reset points from all participants. **Left:** Distance of reset from true path as a single spatial limit. **Right:** Reset coordinates predicted from a single temporal limit (at two speeds)."}
#plotResetPoints()
#plotModels()
plotBootstrappedFits()
```

We test this with a series of simple models, based on straighforward trigonometry. First, we calculate the angle of a line through the origin and a sample on the trajectory halfway to the reset point (with coordinates $X$ and $Y$), as the strength of the illusion ($a = atan2(X,Y)$ expressing the angular deviation the line $x=0$ which is the real path of the gabor). This can then be used to calculate a slope ($s = \frac{sin(a)}{cos(a)}$).

@@

[3a]  &nbsp;&nbsp;&nbsp;&nbsp;&nbsp;  $x = L_t \cdot v \cdot cos(a)$
[3b]  &nbsp;&nbsp;&nbsp;&nbsp;&nbsp;  $y = L_t \cdot v \cdot sin(a)$

Where $L_t$ and $L_x$ are a temporal and spatial limit (in seconds and cm respectively) that are free parameters, $a$ and $s$ corresponds to the calculated angle and slope expressing the strength of the illusion, and speed ($v$ in _cm/s_) is given by the stimulus. If we fit the two simple models for each participant separately, by minimizing the mean square error (error as absolute distance in centimeters between predicted and actual reset points), most participants are best fit with a spatial limit (MSE: 6/9, AIC: 7/9) and the remaining participants are best fit with a temporal limit. We can also combine limits to see if they are complementary. Here, we see that for 4/9 participants (same for MSE and AIC) the best fit boils down to a spatial limit as the other limit is set to 0. Of the remaining participants 4 are best fit with a combined spatial and temporal limit (4/9) and one participant is best fit combining a vertical offset with a temporal limit. This indicates that in this data, a spatial limit explains the reset points best. However even the spatial limit does not explain the data very well.

To obtain population level values for the spatial and temporal limits, we resample reset points from each participant, using the smallest number of detected reset points that fall in the first quadrant, in the group (44), to ensure that each participant contributes the same number of reset points. Both models based on a spatial limit (eq. 2) and on a temporal limit (eq. 3) as well as a model combining these two were fit 1000 times. In the models with two limits there is a strong negative correlation between fitted spatial and temporal limits (r(998)=-0.985, p>.001) indicating parameter redundancy in the combined limit models. That means that the two limits probably explain the same parts of the data, making a two-limit model very hard to interpret. Hence, only the bootstrapped confidence intervals for the two single limit models are shown in Figure 5. Further, since temporal limits might be fit better with an exponential distribution (and minimizing errors amounts to using a normal distribution), we repeated fitting single limit models based on normal, exponential and gamma likelihood distributions for all participants, but found that of the best temporal models, most used a normal distribution (6/9) and the rest an exponential distribution (3/9). All model fits combined show that while the reset point data reported here do somewhat favour a spatial limit, they can not conclusively rule out other limits - most notably a temporal limit.

@@

To measure resets, we asked participants to draw the perceived path of the stimulus after a single pass. The orientation of the drawn traces reflect the strength of the illusion, and in most trials there was a clear reset point, for either a "return" reset or a "hit-the-wall" reset. We then set out to test if these spontaneous resets are triggered by a limiting distance of drift away from the physical location, or by a specific limit of time. The results showed that the resets are better explained by the distance from the true position, not by elapsed time.

Recently, Kwon, Tadin, and Knill [-@Kwon2015] proposed a Bayesian object tracking model to explain the double-drift illusion. In their model, when the position signals are weak, the motion signals bias perceived position away from the true location. In their case, this deviation saturated after about 200 _ms_ and the perceived path then continued parallel to the physical path, as in the “hit the wall” resets we described above [@Kwon2015, Fig. 3]. In contrast to their model, our participants in Experiment 2 showed quite variable timing for the reset point where the accumulation of offset ended. The resets were more stable in terms of spatial offset (Fig. 4, middle) but much more variable in terms of time from the start of the gabor's motion. In any case, the temporal mean of these reset points was not 200 _ms_ but 1.5 _s_ and ranged from 0.5 _s_ to almost 4 _s_ (see Fig. 4). Moreover, 4/9 of our participants regularly reported “return” resets that the Kwon et al. [-@Kwon2015] model does not predict. The “return” resets suggest that both the physical and perceived locations are available to the visual system and that when the deviation exceeds some limit, the perceived location returns toward the physical location. At that point, the build-up of the illusion may start again.

Recently, it has been suggested that resets can be triggered by a distraction of attention [@Nakayama2020]. The participants in this experiment also retraced the perceived path and reported “return” resets. Nakayama and Holcombe also suggest that planned eye movements can trigger a “return” reset back to the physical path. In our experiment, we did not control for eye-movements, so we can't directly investigate whether eye movements accompanied the “return” resets that participants reported. However, “return” resets occurred at a fairly consistent location for each participant, which suggests that they are triggered by the stimulus dynamics and not by attentional distraction or eye movements.

We find that manual tracking of the double-drift stimulus during its motion did show the expected illusion. Few if any resets were detected during the online tracking, either because of the additional attentional requirements or because of the short duration (2 s), Participants did report spontaneous resets in delayed, single-pass re-tracing of Experiment 2. The resets were best explained by a limiting spatial offset from the true stimulus location. This finding suggested that resets resolve a conflict between the physical and illusory position of the stimulus and that both are available to the visual system. This confirms the notion that we gauge the position of objects in the world by combining both retinal position signals and when the retinal information has high uncertainty, motion-based predictions.

>>>>

In our second experiment, we look at the spatial and temporal limits of the illusion. At some point, the illusion stops accumulating. The accumulation can be reset by introducing a temporal break [@Lisi2015] or by distraction attention [@Nakayama2020], but it may also reset spontaneously once the accumulation has gone on too long or too far. Informally, some observers have reported that the gabor’s path may saturate so that it remains parallel to, but offset from the true path. Others report that the position moves back toward its physical location, either slowly or suddenly, whereupon accumulation begins again. The ability to record trajectories of the perceived position will then allow us to study spontaneous reset outcomes.

@@

If $K$ is 0, then there is no illusion and the perceived direction matches the external motion direction. However, if $K$ is 1, then the internal motion has equal contribution with the external motion in determining the perceived direction. For comparison with other work, we will find the value of $K$ that best describes the strength of the illusion in each of our two experiments.

@@

```{r fig2, echo=FALSE, message=FALSE, fig.width=8, fig.height=5, fig.cap="\\label{fig:fig2}**Figure 2:** Online Tracking Data. **Left:** Individual trajectories. Note that there are only a few obvious discontinuities (highlighted in red). In general, the traced paths showed rough linearity. **Top right:** Average heading of the middle 1 _s_ of each pass. Distance from origin is time, angle is the heading angle. Dark blue lines the average heading, in light blue a polar heatmap with the distribution of heading. **Bottom right:** The strength of the illusion in degrees angle deviation from the physical path. In red are the average illusion strengths for 0 _cps_, ±1 _cps_ and ±3 _cps_, and in light blue the data for the 4 individual participants. Gray dashed line: prediction of the vector combination model of Equation 1 with $K=1$. Black line: fit with $K=0.81$ as in Cavanagh & Tse (2019), $K=0.81$. Purple: best fit to our data, $K=0.72$."}
plotBoundedTracking()
```

@@

The individual trajectories reveal some points that are potential spontaneous resets that we try to extract. Using a smooth spline interpolated version of the trajectory we detect right-left direction changes, based on sign changes. Since the initial movement is often jittery, we only take direction changes further than ~0.39 cm ( ~0.23 _dva_ ) away from the starting point. The first (if any) of the remaining direction changes, are then taken as reset points. To keep the horizontal and vertical (straight ahead) coordinates of the reset points isometric they are given in centimeters. The distance from origin can then be converted to time as 1 _cm_ = 3.375 _s_ in the 4 _s_ condition and 1 _cm_ = 4.5 _s_ in the 3 _s_ condition. Participants' ability to retrace perceived trajectories might vary, but normalizing trajectories so that the final Y coordinate is at 13.5 _cm_ does not noticably change the results.

@@

```{r fig4, echo=FALSE, message=FALSE, fig.width=7.5, fig.height=4, fig.cap="\\label{fig:fig4}**Figure 4:** Re-tracing Data. **Left:** Reset points are taken at the first sample in a splined-smoothed re-traced trajectory that goes in the direction opposite that of the expected illusion. Orange and gray lines: example trajectory. Orange dot: reset point. Red dot: point at half the distance to the reset point. Angle a: measure of illusion strength. **Middle:** One participant's trajectories with very clear 'return' resets at the purple dots. (Purple lines: trajectories up to resets, gray lines: trajectories after resets, blue lines: trajectories that had no resets in the re-tracing.) Other participants' data were not this clear. **Right:** Distribution of all 859 detected reset points."}
#plotExampleData()
plotResetPoints()
```

On average, this method detects resets in ~75% of trials. For three participants, only about half the trials showed resets (42%, 44% and 58%). For the remaining six participants the algorithm detected resets in 72%-99% of trials, with an average of 88%. The minimum number of resets in each of the six conditions for any participant is 7 (out of 24 trials), with an average of ~17. Since participants experiencing a "hit-the-wall reset" (Fig. 3) as opposed to a "jump" reset, might not always display a detectable direction change in their paths, 50% of trials with resets seems reasonable. Either type of perceived resets appears valid, and the average reset positions still indicates some limit to the illusion. Therefore, we kept all participants and resets in our analysis (reset points in Fig. 4, right).

As before, we compare the strength of the illusion (see Fig 5, left), with an earlier study [@Cavanagh2019]. A line through the origin with slope 0.51 ($K$ from Eq. 1) predicts illusion strength fairly well in these data, which is weaker than the $K$=0.81 found by Cavanagh & Tse [-@Cavanagh2019] and lower than the $K$=0.72 found in Experiment 1 with online tracking. While the illusion is less strong, it is nevertheless there and depends on the relationship between internal and external speed as in previous studies.

@@

Now we look at the distribution of reset points, and whether they can be explained by a temporal or spatial limit. In Figure 4 (middle), we can see that the reset points vary more along the y-axis, than along the x-axis. That is, it looks like predicting reset points from a single X coordinate (or spatial limit) than predicing reset points from a given time. However, there is also considerable variance, so that fitting any line through the data, will necessarily leave errors.

```{r fig5, echo=FALSE, message=FALSE, fig.width=7.5, fig.height=4, fig.cap="\\label{fig:fig5}**Figure 5:** Illusion strength and reset models. **Left:** Illusion strength as in Figure 2 (bottom right); plotted against predicted strength from Eq.1: the arctangent of the ratio of internal to external velocities. Purple dots are the group means from 3 s passes, red from 4 s passes (2, 3 and 4 _cps_ internal speeds). Predictions from Eq. 1: Gray dashed line for $K=1$, black line; $K=0.81$, purple line; best fit here at $K=0.51$ and the shaded area shows a 95% confidence interval. **Right:** Model limits shown here come from a weighted bootstrapping across participants, lines are median parameter values, shaded areas the 95% confidence intervals of the limits. Gray dots are the reset points from all participants. While a spatial limit provides the best fit, neither a spatial nor a temporal limit explains the data well."}
#plotResetPoints()
#plotModels()
#plotBootstrappedFits()
plotIllusionStrength()
```

We test this with a series of simple models. First, for every reset point (with coordinates $X$ and $Y$) we use the illusion strength, expressed as either an angle (Fig 4, left: $a = atan2(X,Y)$) or a slope ($s = \frac{sin(a)}{cos(a)}$) to predict reset points.

@@

[3a]  &nbsp;&nbsp;&nbsp;&nbsp;&nbsp;  $x = L_t \cdot v \cdot cos(a)$

[3b]  &nbsp;&nbsp;&nbsp;&nbsp;&nbsp;  $y = L_t \cdot v \cdot sin(a)$

Where $L_t$ and $L_x$ are a temporal and spatial limit (in seconds and cm respectively) that are free parameters, $a$ and $s$ corresponds to the calculated angle and slope expressing the strength of the illusion, and speed ($v$ in _cm/s_) is given by the stimulus. If we fit the two simple models for each participant separately, by minimizing the mean square error (in isometric space, with absolute distance in centimeters between predicted and actual reset points), most participants are best fit with a spatial limit (MSE: 6/9, AIC: 7/9) and the remaining participants are best fit with a temporal limit. We can also combine limits to see if they are complementary, including. Here, we see that for 4/9 participants (MSE and AIC) the best fit equals a spatial limit as the other limit is set to 0. Of the remaining participants 4 are best fit with a combined spatial and temporal limit (4/9) and one participant is best fit combining a vertical offset with a temporal limit. This indicates that in this data, a spatial limit explains the reset points best. However even the spatial limit does not explain the data very well.

To obtain population level values for the spatial and temporal limits, we resample reset points from each participant, using the smallest number of detected reset points that fall in the first quadrant, in the group (44), to ensure that each participant contributes the same number of reset points. Both models based on a spatial limit (eq. 2) and on a temporal limit (eq. 3) as well as a model combining these two were fit 1000 times. In the models with two limits there is a strong negative correlation between fitted spatial and temporal limits (r(998)=-0.985, p>.001) indicating parameter redundancy in the combined limit models. That means that the two limits probably explain the same parts of the data, making a two-limit model very hard to interpret. Hence, only the bootstrapped confidence intervals for the two single limit models are shown in Figure 5. Further, since temporal limits might be fit better with an exponential distribution (and minimizing errors amounts to using a normal distribution), we repeated fitting single limit models based on normal, exponential and gamma likelihood distributions for all participants, but found that of the best temporal models, only a few used an exponential distribution (3/9) and the rest a normal distribution (6/9). All model fits combined show that while the reset point data reported here do somewhat favour a spatial limit, they can not conclusively rule out other limits - most notably a temporal limit.

@@

To measure resets, we asked participants to draw the perceived path of the stimulus after a single pass. The orientation of the drawn traces reflect the strength of the illusion, and in most trials there was a clear reset point, for either a "return" reset or a "hit-the-wall" reset. We then set out to test if these spontaneous resets are triggered by a limiting distance of drift away from the physical location, or by a specific limit of time. The results showed that the resets are slightly better explained by the distance from the true position, but a temporal limit is not ruled out.

Recently, Kwon, Tadin, and Knill [-@Kwon2015] proposed a Bayesian object tracking model to explain the double-drift illusion. In their model, when the position signals are weak, the motion signals bias perceived position away from the true location. In their case, this deviation saturated after about 200 _ms_ and the perceived path then continued parallel to the physical path, as in the “hit the wall” resets we described above [@Kwon2015, Fig. 3]. In contrast to their model, our participants in Experiment 2 showed quite variable timing for the reset point where the accumulation of offset ended. The resets were more stable in terms of spatial offset (Fig. 4, middle) but much more variable in terms of time from the start of the gabor's motion. In any case, the temporal limit of these reset points was not 200 _ms_ but 1.1 _s_ and fell into a wide range from 0.5 _s_ to almost 4 _s_ across trials (see Fig. 4). Moreover, 4/9 of our participants regularly reported “return” resets that the Kwon et al. [-@Kwon2015] model does not predict. The “return” resets suggest that both the physical and perceived locations are available to the visual system and that when the deviation exceeds some limit, the perceived location returns toward the physical location. At that point, the build-up of the illusion may start again.

Recently, it has been suggested that resets can be triggered by a distraction of attention [@Nakayama2020]. The participants in this experiment also retraced the perceived path and reported “return” resets. Nakayama and Holcombe also suggest that planned eye movements can trigger a “return” reset back to the physical path. In our experiment, we did not control for eye-movements, so we can't directly investigate whether eye movements accompanied the “return” resets that participants reported. **However, “return” resets occurred at a fairly consistent location for each participant, which suggests that they are triggered by the stimulus dynamics and not by attentional distraction or eye movements.**

We find that manual tracking of the double-drift stimulus during its motion did show the expected illusion. Few if any resets were detected during the online tracking, either because of the additional attentional requirements or because of the short duration (2 s), Participants did report spontaneous resets in delayed, single-pass re-tracing of Experiment 2. The resets were best explained by a limiting spatial offset from the true stimulus location, but do not exclude a temporal limit. These findings may either indicate there are multiple causes for spontaneous resets or that a re-tracing method is not suitable to resolve what causes this phenomenon. Either way, both a random time from stimulus onset until reset and too large a conflict between retinal and perceived position leading to resets remain possible.
